---
title: "Assembling a New York City Council Discretionary Funding Dataset"
subtitle: "FY 2006 - 2026"
format: 
  html:
    code-fold: true
---

```{r}
#| label: setup
#| include: true
#| warning: false
#| message: false

# Notebook setup

# Use quarto render to generate a static site in /docs
# Set github pages to deploy from main branch, /docs

# Insure packages are installed by checking required packages against installed packages
reqd_pkgs <- c("flextable",
               "sf",
               "tidyverse",
               "mapgl",
               "htmltools",
               "officer",
               "forcats",
               "socratadata",
               "mapgl",
               "mapview",
               "DBI",
               "RPostgres",
               "httr2",
               "rvest",
               "readxl")

install.packages(
  setdiff(
    reqd_pkgs,
    rownames(installed.packages())
  ),
  repos = "https://cloud.r-project.org"
)

lapply(reqd_pkgs, library, character.only = TRUE) |> 
  invisible()

# If needed, ensure that the Quarto render and RStudio read the same environment vars
readRenviron(".Renviron")

### UNCOMMENT FOR POSTGIS VIA DOCKER ###########################################

# # Start Postgres via Docker, wait for healthcheck
# system("docker compose up -d --wait")
# 
# # Set up connection from env
# con <- dbConnect(RPostgres::Postgres())
# 
# # Enable PostGIS
# invisible(dbExecute(con, "CREATE EXTENSION IF NOT EXISTS postgis"))

###############################################################################
# Get cil_table()
source("cil_table.R")
```

## Data sources

The chunks in this section download each source as published into `data/raw/`. By default, they do not run when the notebook is rendered. Download them by hand by setting the `eval` chunk option to `true`. Each download overwrites the files in `data/raw/` and rewrites `data/raw/manifest.csv`.

```{r}
#| label: raw-helpers
#| eval: true

raw_dir <- "data/raw"

# One manifest row per downloaded file. The Council reissues its workbooks
# under the same file name, so the download time and checksum are what tell
# two copies of a fiscal year apart.
manifest_row <- function(source, item, path, url, source_modified = NA, rows = NA) {
  tibble(
    source          = source,
    item            = as.character(item),
    path            = path,
    url             = url,
    fetched_at      = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    source_modified = as.character(source_modified),
    bytes           = file.size(path),
    rows            = as.integer(rows),
    sha256          = unname(tools::sha256sum(path))
  )
}

# Saves a file byte-for-byte and returns the server's Last-Modified date.
download_raw <- function(url, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  modified <- request(url) |>
    req_retry(max_tries = 3) |>
    req_perform(path = path) |>
    resp_header("last-modified")
  if (is.null(modified)) return(NA)
  format(curl::parse_date(modified), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
}

cil_table(
  tribble(
    ~`Manifest Column`, ~Description,
    "source", "Source file type: open_data, workbook or schedule_c.",
    "item", "The dataset ID for Open Data. The fiscal year for workbooks and Schedule C.",
    "path", "Where the file was saved, relative to the project root.",
    "url", "The address the file was downloaded from.",
    "fetched_at", "When the file was downloaded, in UTC.",
    "source_modified", "When the publisher last changed the file, in UTC. For Open Data, the dataset's last data update. For workbooks and Schedule C, the server's Last-Modified header.",
    "bytes", "Size of the saved file in bytes.",
    "rows", "Number of rows in the table. Recorded for Open Data only.",
    "sha256", "SHA-256 checksum of the saved file."
  )
)
```

### NYC Open Data

Four tables, read with `{socratadata}` and written to CSV.

```{r}
#| label: open-data-table
#| echo: false

cil_table(
  tribble(
    ~Dataset, ~ID,
    "City Council Discretionary Funding", "4d7f-74pe",
    "City Council Members, 1999 to Present", "uvw5-9znb",
    "Discretionary Award Tracker", "ujre-m2tj",
    "Recent Contract Awards", "qyyg-4tf5"
  ),
  full_height = TRUE
)
```

```{r}
#| label: raw-open-data
#| eval: false

open_data <- c(
  council_discretionary_funding = "4d7f-74pe",
  council_members               = "uvw5-9znb",
  discretionary_award_tracker   = "ujre-m2tj",
  recent_contract_awards        = "qyyg-4tf5"
)

manifest_open_data <- imap(open_data, \(id, name) {
  url  <- paste0("https://data.cityofnewyork.us/d/", id)
  path <- file.path(raw_dir, "open_data", paste0(name, "_", id, ".csv"))
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  # alias = "drop" keeps the API field names (mocs_id, council_member) instead
  # of the display labels.
  data <- soc_read(url, alias = "drop")
  updated <- format(attr(data, "data_last_updated"), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
  # A Socrata url column arrives as a nested url/description pair. The portal's
  # own CSV export writes the url alone, so that is what is written here.
  data <- mutate(data, across(where(is.data.frame), \(x) x$url))
  # The portal publishes dates as midnight with no time zone. Written as the
  # date alone, so a UTC conversion cannot move them.
  data <- mutate(data, across(where(is.POSIXct), as_date))
  write_csv(data, path, na = "")
  manifest_row("open_data", id, path, url, updated, nrow(data))
}) |>
  list_rbind()
```

### Disclosure workbooks

`funded_disclosure_FY####`, one Excel file per fiscal year, FY2009 through FY2026, saved as published. These are the files the Council's [expense funding search](https://rnd.council.nyc.gov/expense_funding/) loads.

```{r}
#| label: raw-workbooks
#| eval: false

manifest_workbooks <- map(2009:2026, \(fy) {
  file <- paste0("funded_disclosure_FY", fy, if (fy <= 2013) ".xls" else ".xlsx")
  url  <- paste0("https://rnd.council.nyc.gov/src/assets/excelFund/", file)
  path <- file.path(raw_dir, "workbooks", file)
  modified <- download_raw(url, path)
  manifest_row("workbook", fy, path, url, modified)
}) |>
  list_rbind()
```

### Schedule C

The Adopted Expense Budget Adjustment Summary, FY2008 through FY2027. There is no index of these PDFs: each is linked from its fiscal year's page at `council.nyc.gov/budget/fy####/`. The FY2011 page links two documents, a Schedule C and the Changes Pursuant to Section 254; both are downloaded.

```{r}
#| label: raw-schedule-c
#| eval: false

manifest_schedule_c <- map(2008:2027, \(fy) {
  page  <- paste0("https://council.nyc.gov/budget/fy", fy, "/")
  links <- read_html(page) |> html_elements("a")
  links <- tibble(
    text = html_text2(links),
    url  = xml2::url_absolute(html_attr(links, "href"), page)
  ) |>
    filter(str_detect(url, regex("\\.pdf$", ignore_case = TRUE)))

  schedule_c <- links |>
    filter(str_detect(text, regex("schedule c", ignore_case = TRUE))) |>
    slice(1) |>
    mutate(file = paste0("Schedule_C_FY", fy, ".pdf"))
  # From FY2020 on, "Section 254" on these pages titles capital budget
  # documents, so it is only followed for FY2011.
  section_254 <- links |>
    filter(fy == 2011, str_detect(text, regex("section 254", ignore_case = TRUE))) |>
    slice(1) |>
    mutate(file = paste0("Schedule_C_FY", fy, "_Section_254.pdf"))

  if (nrow(schedule_c) == 0) stop("No Schedule C link on ", page, call. = FALSE)

  bind_rows(schedule_c, section_254) |>
    pmap(\(text, url, file) {
      path <- file.path(raw_dir, "schedule_c", file)
      modified <- download_raw(url, path)
      manifest_row("schedule_c", fy, path, url, modified)
    }) |>
    list_rbind()
}) |>
  list_rbind()
```

### Manifest

```{r}
#| label: raw-manifest
#| eval: false

bind_rows(manifest_open_data, manifest_workbooks, manifest_schedule_c) |>
  write_csv(file.path(raw_dir, "manifest.csv"))
```

### Reference tables

The Institute's reference tables are read from `capi-docs/Council_Discretionary_Reference_Tables.xlsx`. They are not downloaded.

## Build

### Reading the disclosure workbooks

The eighteen workbooks are read into one table. Each file's first tab is read with every column as text. Nothing is cleaned at this stage.

Column names change from year to year. Each published header is renamed to match the fields used in the NYC Open Data datasets.

```{r}
#| label: workbook-crosswalk

workbook_crosswalk <- tribble(
  ~open_data_name,              ~workbook_header,
  "mocs_id",                    "MOCS ID#",
  "fiscal_year",                "Fiscal Year",
  "source",                     "Source",
  "council_member",             "Council Member",
  "council_member",             "Council Members",
  "legal_name_of_organization", "Legal Name",
  "legal_name_of_organization", "Legal Name of Organization",
  "legal_name_of_organization", "Legal Name of Organization Requesting Funding",
  "ein",                        "EIN",
  "ein",                        "Tax ID",
  "status",                     "Status",
  "amount",                     "Amount",
  "amount",                     "Amount ($)",
  "amount",                     "Amount ($",
  "agency",                     "Agency",
  "program_name",               "Program Name",
  "address",                    "Street Address 1",
  "address",                    "Address",
  "address",                    "Grantee Street Address 1",
  "address_2_optional",         "Street Address 2",
  "address_2_optional",         "Address 2 (optional)",
  "address_2_optional",         "Address 2",
  "address_2_optional",         "Address 2 (optional",
  "address_2_optional",         "Grantee Street Address 2",
  "city",                       "City",
  "state",                      "State",
  "postcode",                   "Zip Code",
  "postcode",                   "Postal Code",
  "purpose_of_funds",           "Purpose of Funds",
  "purpose_of_funds",           "Purpose to be listed in Schedule C",
  "purpose_of_funds",           "CM Purpose of Funds",
  "fiscal_conduit_name",        "Fiscal Conduit Name",
  "fiscal_conduit_name",        "Fiscal Conduit",
  "fc_ein",                     "FC EIN",
  "fc_ein",                     "Fiscal Conduit EIN"
)

workbook_paths <- list.files(file.path(raw_dir, "workbooks"), full.names = TRUE)
workbook_fy    <- as.integer(str_extract(basename(workbook_paths), "\\d{4}"))
```

```{r}
#| label: workbook-crosswalk-table
#| echo: false

# Collapses a set of fiscal years into ranges: FY2009–2011, FY2013.
fy_ranges <- function(fy) {
  fy   <- sort(unique(fy))
  runs <- split(fy, cumsum(c(1, diff(fy) != 1)))
  map_chr(runs, \(run) {
    if (length(run) == 1) paste0("FY", run) else paste0("FY", min(run), "–", max(run))
  }) |>
    paste(collapse = ", ")
}

# The years come from the headers in the files themselves.
workbook_headers <- map2(workbook_paths, workbook_fy, \(path, fy) {
  tibble(fiscal_year = fy, workbook_header = names(read_excel(path, sheet = 1, n_max = 0)))
}) |>
  list_rbind()

workbook_crosswalk |>
  left_join(workbook_headers, by = "workbook_header") |>
  summarise(fiscal_years = fy_ranges(fiscal_year), .by = c(open_data_name, workbook_header)) |>
  rename(
    `Open Data name`  = open_data_name,
    `Workbook header` = workbook_header,
    `Fiscal years`    = fiscal_years
  ) |>
  cil_table(full_height = TRUE)
```

Seven workbooks, FY2009 through FY2014 and FY2016, have no fiscal year column. For those the year is taken from the file name.

The FY2024 workbook has seventeen rows below its last award: twelve empty rows and five rows of totals. Rows with neither a source nor a legal name are dropped on read. No other workbook has rows meeting this condition.

```{r}
#| label: read-workbooks

is_blank <- function(x) is.na(x) | str_trim(x) == ""

read_workbook <- function(path, fy) {
  # trim_ws = FALSE keeps leading and trailing spaces in the values as published.
  data <- read_excel(path, sheet = 1, col_types = "text", trim_ws = FALSE)

  unmapped <- setdiff(names(data), workbook_crosswalk$workbook_header)
  if (length(unmapped) > 0) {
    stop(basename(path), " has a header the crosswalk does not cover: ",
         paste(unmapped, collapse = ", "), call. = FALSE)
  }
  names(data) <- workbook_crosswalk$open_data_name[match(names(data), workbook_crosswalk$workbook_header)]

  if (!"fiscal_year" %in% names(data)) data$fiscal_year <- as.character(fy)

  filter(data, !(is_blank(source) & is_blank(legal_name_of_organization)))
}

workbooks <- map2(workbook_paths, workbook_fy, read_workbook) |>
  list_rbind() |>
  select(all_of(unique(workbook_crosswalk$open_data_name)))
```

### Reading the Open Data table

City Council Discretionary Funding (4d7f-74pe) is read from the downloaded CSV with every column as text, as the workbooks are.

```{r}
#| label: read-open-data

open_data_funding <- read_csv(
  file.path(raw_dir, "open_data", "council_discretionary_funding_4d7f-74pe.csv"),
  col_types = cols(.default = col_character()),
  na = ""
)
```

### Comparing the workbooks with Open Data

The two sources cover the same designations for FY2009 through FY2021. A workbook row and an Open Data row are counted as the same award when fiscal year, source, organization name, and amount all agree. Before comparing, names are lowercased, punctuation is removed, and repeated spaces are collapsed; sources are lowercased and their spaces collapsed; amounts are written to two decimal places. Where a key appears more than once in a year, rows are paired off one for one.

```{r}
#| label: compare-sources

match_keys <- function(data) {
  data |>
    transmute(
      fiscal_year = as.integer(fiscal_year),
      amount      = as.numeric(amount),
      key = paste(
        str_squish(str_to_lower(source)),
        str_squish(str_remove_all(str_to_lower(legal_name_of_organization), "[[:punct:]]")),
        sprintf("%.2f", amount),
        sep = "|"
      )
    )
}

workbook_keys  <- match_keys(workbooks)
open_data_keys <- match_keys(open_data_funding)

source_totals <- full_join(
  summarise(workbook_keys, workbook_rows = n(), workbook_amount = sum(amount), .by = fiscal_year),
  summarise(open_data_keys, open_data_rows = n(), open_data_amount = sum(amount), .by = fiscal_year),
  by = "fiscal_year"
)

source_matches <- full_join(
  count(workbook_keys, fiscal_year, key, name = "in_workbook"),
  count(open_data_keys, fiscal_year, key, name = "in_open_data"),
  by = c("fiscal_year", "key")
) |>
  replace_na(list(in_workbook = 0, in_open_data = 0)) |>
  summarise(
    matched        = sum(pmin(in_workbook, in_open_data)),
    workbook_only  = sum(pmax(in_workbook - in_open_data, 0)),
    open_data_only = sum(pmax(in_open_data - in_workbook, 0)),
    .by = fiscal_year
  )

source_comparison <- source_totals |>
  left_join(source_matches, by = "fiscal_year") |>
  arrange(fiscal_year) |>
  # Open Data stops at FY2021, so there is nothing to match against after it.
  mutate(across(c(matched, workbook_only, open_data_only), \(x) if_else(is.na(open_data_rows), NA, x)))

source_comparison |>
  transmute(
    `Fiscal year`    = paste0("FY", fiscal_year),
    `Workbook rows`  = workbook_rows,
    `Open Data rows` = open_data_rows,
    Matched          = matched,
    `Workbook only`  = workbook_only,
    `Open Data only` = open_data_only,
    `Workbook $ minus Open Data $` = round(workbook_amount - open_data_amount)
  ) |>
  mutate(across(-`Fiscal year`, \(x) if_else(is.na(x), "", format(x, big.mark = ",", trim = TRUE)))) |>
  cil_table(full_height = TRUE)
```
